---
title: "The Fabric of Interactive Graphics"
author: "Adam Bartonicek"
format: 
  revealjs:
    theme: [default, styles.scss]
editor: visual
editor_options: 
  chunk_output_type: console
execute:
  fig-dpi: 150
bibliography: references.bib
---

## Thesis topic evolution

- Master's thesis: implement a portable interactive data visualization library in JS for R, building on `iplots` [@urbanek2011]
- Wanted to incorporate Grammar of Graphics [GoG, @wilkinson2012]
- Also had interest in functional programming at the time
  - Realized there are some important parallels
- This lead me to category theory [@fong2019]

## Historical split in interactive graphics

-   **"Classical" IDV systems (1980-2010s)**
    - Standalone apps
    - Deep interactivity, less flexible
    - Nominal style: *scatterplot, barplot, ...*
-   **Modern web-based IDV systems (2005-current)** 
    - Highly customizable but less interactivity out-of-the-box
    - Compositional style: *plot with continuous x- and y- encodings and point geometry ...* (GoG-based)

## Why is it challenging to combine the GoG modularity with interactivivity? {.center}

## Graphics and structure

-   A plot is not just a picture - instead, it is a graphical representation of structured information

## Plots as maps

- A plot is a map (function) from data to a graphical display (represented on computer screen/paper):

$$\text{Data} \to \text{Graphics}$$

- *Can this map be any old function?*
  - Plots can be misleading representations [@tufte2001; @cairo2014]
  - Faithful plots "in the wild" exhibit many regular features
  - Interactivity adds additional constraints

## Plots as partitions

- Useful idea: geoms as sets, plots as *partitions*:
    -   A surjective collection of disjoint subsets
    -   *= Each point, bar, or line represents a distinct set of cases/rows, and together the objects cover the whole data space (bijection)*

$$\text{Data subsets} \to \text{Geometric objects}$$
    
-   Some exceptions:
    -   2D KDE plots (multiple isopleths map to the same datapoints)
    -   Graphical deficiencies (e.g. overplotting)

## Plots as partitions and interactivity

- The (bijective) partition model is well-suited for interactive graphics
- Provides unambiguous visual mapping in linked selection:
  - *If I click a bar in a plot, only that bar gets highlighted* <br> (not the case with non-disjoint data subsets)
  
```{r}
#| echo: false
#| fig-width: 30
#| fig-align: "center"
library(ggplot2)
library(patchwork)

theme_set(
  theme_bw() + theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks = element_blank(),
    panel.border = element_blank()
  ))

mtcars$am <- factor(mtcars$am)

p1 <- ggplot(mtcars, aes(cyl, mpg)) +
  geom_bar(stat = "summary", fun = "sum", fill = "grey80") +
  geom_bar(data=subset(mtcars, cyl == 4),
           stat = "summary", fun = "sum", fill = "grey80", 
           col="grey20", linetype="dashed", width = 1.8, size=1.5) +
  annotate("segment", x=3.9, y=230, xend=4.1, yend=210, 
           arrow=arrow(type = "closed", length = unit(0.1, "in")),
           lwd=1.5, linejoin="mitre") +
  guides(linetype="none")

p12 <- ggplot(mtcars, aes(cyl, mpg)) +
  geom_bar(stat = "summary", fun = "sum", alpha=0) +
  annotate("text", x = 6, y = 150, label = "vs.", size=14)

p2 <- ggplot(mtcars, aes(cyl, mpg, fill = cyl == 4)) +
  geom_bar(stat = "summary", fun = "sum") +
  scale_fill_manual(values=c("grey80", "steelblue")) +
  guides(fill = "none") 

mtcars$highlight <- mtcars$am == "1" | mtcars$cyl == 4

p3 <- ggplot(mtcars, aes(cyl, mpg, fill = highlight)) +
  geom_bar(stat = "summary", fun = "sum") +
  scale_fill_manual(values=c("grey80", "steelblue")) +
  guides(fill = "none") 

p1 + plot_spacer() + p2 + p12 + p1 + plot_spacer() + p3 + 
  plot_layout(nrow=1, widths = c(1, 0.2, 1, 1, 1, 0.2, 1))
```
  
- Also useful for querying: geom/subset information is independent

## Plots, summaries, and aesthetics

- We rarely plot raw data: instead, we often summarize/aggregate
- Also encode the summaries into aesthetics (scales/coord. systems)
- Therefore, the full mapping is a *composition* of several maps:

$$\text{Data subsets} \to \text{Summaries} \to \text{Aesthetics} \to \text{Geometric objects}$$

- *The domains/codomains of these maps often have structure which the maps may or may not preserve*

## Example: Stacking 1

- The data partition can have a hierarchical structure
  - = Data subsets made up of smaller subsets
  - E.g. stacked bar made up of segments, pie chart
  - *Not purely visual* - segments represent data information
- The map should preserve this underlying structure
  - I.e. aggregation and encoding also need to be structure-preserving

## Example: Stacking 2

- Suppose: 

::: {style="margin-top: -30px; margin-bottom: -30px;"}
$$\qquad \text{Bar subset} = \text{Gray subset} \cup \text{Blue subset}$$
:::

- We want: 

::: {style="margin-top: -30px; margin-bottom: -30px;"}
$$\qquad \text{Bar} = \text{Stack}(\text{Gray segment}, \text{Blue segment})$$
:::

```{r}
#| echo: false
#| fig-height: 1
#| fig-width: 1.5
#| fig-align: "center"

ggplot(mtcars, aes(cyl, mpg, fill = am)) +
  geom_bar(stat = "summary", fun = "sum") +
  scale_fill_manual(values=c("grey80", "steelblue")) +
  guides(fill = "none") 
```

## Example: Stacking 3

- The mapping should preserve salient visual properties
- E.g. bar height:

::: {style="margin-top: -15px; margin-bottom: -15px; font-size: 90%;"}
$$\text{Height(Bar subset)} = \text{Height(Stack(Blue subset, Red subset))}$$
:::

- Not always the case:
  - *"Stacking bars that represent counts, sums, or percentages is fine, but a stacked bar chart where bars show average values is generally meaningless." [@wills2011, p. 112]*
  - *Why* is it meaningless? And *when*?  

## Category theory and functors

- Category theory is an area of math concerned with structure and composition [@fong2019]
- Fundamental concept: a *functor*
- Functors are a generalization of structure-preserving maps
- Specifically, functors preserve function composition:

$$F(f ; g) = F(f) ; F(g)$$

## Functors and visualization

- In our case, we can think of $f$ and $g$ as union with some fixed set
  - E.g. $f(x) = x \cup A$, and $g(x) = x \cup B$
  - $x$ can be $\varnothing$
- $F$ is the aggregation (and encoding) map. Then:

::: {style="margin-top: -15px; margin-bottom: -15px; font-size: 90%;"}
$$F(f ; g) = F(f) ; F(g)$$
:::

- Translates to e.g.:

::: {style="margin-top: -15px; margin-bottom: -15px; font-size: 90%;"}
$$\text{stack}(\text{sum}(A \cup B)) = \text{stack}(\text{sum}(A), \text{sum}(B))$$
:::

## Functorial summaries and monoids

- Original idea: model plots as *partitions* 
  - Geoms as disjoint sets
- If we start from this, we can make use of *monoid*s
- A monoid $(M, e, \otimes)$ is a set $M$ equipped with a neutral element $e$ and an associative binary operation $\otimes$ 
- It is easy to show that aggregating with a monoid is a functor from a partition category to a category of summaries (preserves set union)
  - Known in parallel computing and relational database research [@parent2018; @fegaras2017; @lin2013; @gibbons2018]

## Preserving properties

-   However, perhaps we want plots to be more than just *any* function
-   Maps can preserve or fail to preserve certain properties
-   If partition is the fundamental model, then the mapping (plot) should preserve the properties of sets and set operations
-   Statistics and graphics should behave like sets

## Set union and associativity

-   One fundamental thing we can do with sets is to merge them (union)
-   Set union has several important properties, including associativity:

$$A \cup (B \cup C) = (A \cup B) \cup C = A \cup B \cup C$$

-   *"If we gather up the stuff contained in several sets, we end up with the same result regardless of what order we do it in"*

## Preserving set union

-   If we want to interpret the geometric objects in our plots as sets, they need to preserve associativity
-   To do this, not just the geometric objects but also the underlying summaries need to be associative
-   Counterexample: stacking

## References
