---
title: "The Fabric of Interactive Graphics"
author: "Adam Bartonicek"
format: 
  revealjs:
    theme: [default, styles.scss]
editor: visual
editor_options: 
  chunk_output_type: console
execute:
  fig-dpi: 150
---

## 

## Interaction vs. modularity

-   "Classical" IDV systems: deep interactivity but less flexible
-   Modern GoG-based IDV systems: highly customizable but usually only shallow interactive features out-of-the-box\
-   Nominal vs. composition-based approach to graphics:
    -   *Scatterplot* vs.
    -   *Plot with continuous x- and y- encodings and point geometry*

## Why is it challenging to combine the GoG modularity with interactivivity?

## Graphics and structure

-   A plot is not just a picture - instead, it is a graphical representation of structured information

## Plots as maps

- A plot is a map (function) from data to a graphical display (represented on computer screen/paper):

$$\text{Data} \to \text{Graphics}$$

- *Can this map be any old function?*
  - Plots "in the wild" exhibit a number of regular properties/features
  - Many in static plots, interactivity adds additional constraints

## Plots as partitions

- Useful to think about plots as *partitions*:
    -   A surjective collection of disjoint (and non-empty) subsets
    -   = *Each point, bar, or line represents a distinct set of cases/rows, and together the objects cover the whole data space (bijection)*

$$\text{Data subsets} \to \text{Geometric objects}$$
    
-   Some exceptions:
    -   2D KDE plots (multiple isopleths map to same datapoints)
    -   Graphical deficiencies (e.g. overplotting)

## Plots as partitions and interactivity

- The (bijective) partition model is useful in interactive graphics
- Provides unambiguous visual mapping in linked selection:
  - *If I click a bar in a plot, only that bar gets highlighted* <br> (not the case with non-disjoint data subsets)
  
```{r}
#| echo: false
#| fig-width: 30
#| fig-align: "center"
library(ggplot2)
library(patchwork)

theme_set(
  theme_bw() + theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    axis.ticks = element_blank(),
    panel.border = element_blank()
  ))

mtcars$am <- factor(mtcars$am)

p1 <- ggplot(mtcars, aes(cyl, mpg)) +
  geom_bar(stat = "summary", fun = "sum", fill = "grey80") +
  geom_bar(data=subset(mtcars, cyl == 4),
           stat = "summary", fun = "sum", fill = "grey80", 
           col="grey20", linetype="dashed", width = 1.8, size=1.5) +
  annotate("segment", x=3.9, y=230, xend=4.1, yend=210, 
           arrow=arrow(type = "closed", length = unit(0.1, "in")),
           lwd=1.5, linejoin="mitre") +
  guides(linetype="none")

p12 <- ggplot(mtcars, aes(cyl, mpg)) +
  geom_bar(stat = "summary", fun = "sum", alpha=0) +
  annotate("text", x = 6, y = 150, label = "vs.", size=14)

p2 <- ggplot(mtcars, aes(cyl, mpg, fill = cyl == 4)) +
  geom_bar(stat = "summary", fun = "sum") +
  scale_fill_manual(values=c("grey80", "steelblue")) +
  guides(fill = "none") 

mtcars$highlight <- mtcars$am == "1" | mtcars$cyl == 4

p3 <- ggplot(mtcars, aes(cyl, mpg, fill = highlight)) +
  geom_bar(stat = "summary", fun = "sum") +
  scale_fill_manual(values=c("grey80", "steelblue")) +
  guides(fill = "none") 

p1 + plot_spacer() + p2 + p12 + p1 + plot_spacer() + p3 + 
  plot_layout(nrow=1, widths = c(1, 0.2, 1, 1, 1, 0.2, 1))
```
  
- Also useful for querying: geom/subset information is independent

## Plots, summaries, and aesthetics

- We rarely plot raw data - instead, we often summarize/aggregate
- Also, we encode the summaries into aesthetics <br> (via scales/coordinate systems)
- Therefore, the full mapping is a *composition* of several maps:

$$\text{Data subsets} \to \text{Summaries} \to \text{Aesthetics} \to \text{Geometric objects}$$

- *The domains/codomains often have structure which the maps may or may not preserve*

## Example: Stacking 1

- The data partition can have a hierarchical structure
  - Data subsets are made up of smaller data subsets
- Example: stacked barplot made up of segments
- To be a valid representation, the visualization map (plot) should preserve this structure
  - I.e. aggregation and encoding also need to be structure-preserving

## Example: Stacking 2 {}

- Suppose: 

::: {style="margin-top: -40px; margin-bottom: -40px;"}
$$\qquad \text{Bar subset} = \text{Gray subset} \cup \text{Blue subset}$$
:::

- We want: 

::: {style="margin-top: -40px; margin-bottom: -40px;"}
$$\qquad \text{Bar} = \text{Stack}(\text{Gray segment}, \text{Blue segment})$$
:::

```{r}
#| echo: false
#| fig-height: 1
#| fig-width: 1.5
#| fig-align: "center"

ggplot(mtcars, aes(cyl, mpg, fill = am)) +
  geom_bar(stat = "summary", fun = "sum") +
  scale_fill_manual(values=c("grey80", "steelblue")) +
  guides(fill = "none") 
```

## Example: Stacking 3

- However, the mapping should preserve salient visual properties
- E.g. bar height:

::: {style="margin-top: -30px; margin-bottom: -30px; font-size: 90%;"}
$$\text{Height(Bar)} = \text{Height(Stack(Blue segment, Red segment))}$$
:::

- However, this is not always the case
  - Stacking sums is fine, stacking other is meaningless

## Structure-preserving maps: Functors

- In category theory, structure-preserving maps are known as functors
- Specifically, functors preserve function composition:

$$F(f ; g) = F(f) ; F(g)$$

## Functors and visualization

$$F(f ; g) = F(f) ; F(g)$$

- In our case, we can think of $f$ and $g$ as taking a union with some fixed set, e.g. $f(x) = x \cup A$ and $g(x) = x \cup B$ (where $x$ can be $\varnothing$)
- $F$ can be the aggregation and encoding maps
- When both aggregation and encoding are functors, then the whole visualization map (plot) is a functor as well (functors compose):

$$G(F(f ; g)) = G(F(f)) ; G(F(g))$$

## Reasoning about graphics

- Practically, functors allow us to reason about composition of summary statistics and graphics
- For instance, sums and maximums compose:
  - $\text{sum}(A \cup B \cup C) = \text{sum}(A) + \text{sum}(B) + \text{sum}(C)$
  - $\text{max}(A \cup B \cup C) = \text{max}(\{\text{max}(A), \text{max}(B), \text{max}(C)\})$
- However, e.g. means do not:
  - $\text{mean}(A \cup B \cup C) \neq \text{mean}(\{\text{mean}(A), \text{mean}(B), \text{mean}(C)\})$
- 

## Preserving properties

-   However, perhaps we want plots to be more than just *any* function
-   Maps can preserve or fail to preserve certain properties
-   If partition is the fundamental model, then the mapping (plot) should preserve the properties of sets and set operations
-   Statistics and graphics should behave like sets

## Set union and associativity

-   One fundamental thing we can do with sets is to merge them (union)
-   Set union has several important properties, including associativity:

$$A \cup (B \cup C) = (A \cup B) \cup C = A \cup B \cup C$$

-   *"If we gather up the stuff contained in several sets, we end up with the same result regardless of what order we do it in"*

## Preserving set union

-   If we want to interpret the geometric objects in our plots as sets, they need to preserve associativity
-   To do this, not just the geometric objects but also the underlying summaries need to be associative
-   Counterexample: stacking
